---
title: "Evans Cohort Secondary Analysis with EpiStats"
author: "Boaz Baliejukia"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


This is a secondary analysis of the Multiparameter Intelligent Monitoring in Intensive Care II
database, consisting o 24,000 patients admitted to the Beth Israel Deaconess Medical Center
ICU between 2001 and 2008. Patients requiring mechanical ventilation who did not require vasopressors
or have a diagnosis of sepsis were identified (1776 patients), and the primary outcome was 28-day mortality. Our objective was to investigate the **association between the use of indwelling arterial catheters (IAC) and mortality in hemodynamically stable patients with respiratory failure**, using several methods like multivariate regression and stratification to adjust for confounding.

## Loading useful packages

```{r, results='hide', message=FALSE}
library(tidyverse) # for data handling
library(EpiStats)  # for epidemiology
library(knitr)      # to get well formatted tables
library(rstatix)
library(gtsummary)
library(report)
library(finalfit)
```

## load in data

Let's load in the dataset from the project directory using the read.csv() function and call it catheter. The dataset has 46 variables and 1776 selected participants who met the inclusion criteria.
```{r}
catheter <- read.csv("full_cohort_data.csv")

glimpse(catheter)
```

## Data handling

Let's select and rename variables of interest: the exposure (Use of IAC), the outcome (death) and potential confounders (age; sex; sofa score; service unit; coincident diseases like COPD and other respiratory diseases, congestive heart failure (CHF), Atrial fibrillation (AF), chronic renal disease, chronic liver diseases, coronary artery disease, stroke and malignancy; and laboratory tests such as wbc, hemoglobin, platelet, sodium, potassium, bicarbonate, chloride, urea, creatinine, po2 and pco2 .

lack of pneumonia, DNR at admission and Change in code status during ICU admission.
```{r}
variables.of.int <- catheter %>% 
  select(IAC = aline_flg, 
         death = day_28_flg, 
         age, 
         sex = gender_num, 
         sofa = sofa_first, 
         service_unit, 
         COPD = copd_flg,
         Other.resp.D = resp_flg, 
         CHF = chf_flg, 
         AF = afib_flg, 
         chr.renal.D = renal_flg, 
         chr.liver.D = liver_flg,
         coronary.D = cad_flg, 
         stroke = stroke_flg, 
         malignancy = mal_flg, 
         wbc = wbc_first, 
         hemoglobin = hgb_first, 
         platelet = platelet_first,
         sodium = sodium_first, 
         potassium = potassium_first,
         bicarbonate = tco2_first, 
         chloride = chloride_first,
         urea = bun_first, 
         creat = creatinine_first, 
         po2 = po2_first,
         pco2 = pco2_first)

head(variables.of.int)
```
Let's recode categorical variables to get summary statistics table

```{r}
variables.of.int.recoded <- variables.of.int %>% 
  mutate(IAC = case_when(IAC == 1 ~ "yes" , IAC == 0 ~ "no" )) %>% 
  mutate(death = case_when(death == 1 ~ "yes" , death == 0 ~ "no" )) %>%
  mutate(sex = case_when(sex == 1 ~ "male" , sex == 0 ~ "female" )) %>% 
  mutate(COPD = case_when(COPD == 1 ~ "yes" , COPD == 0 ~ "no" )) %>%
  mutate(Other.resp.D = case_when(Other.resp.D == 1 ~ "yes" , Other.resp.D == 0 ~ "no" )) %>%
  mutate(CHF = case_when(CHF == 1 ~ "yes" , CHF == 0 ~ "no" )) %>%
  mutate(AF = case_when(AF == 1 ~ "yes" , AF == 0 ~ "no" )) %>%
  mutate(chr.renal.D = case_when(chr.renal.D == 1 ~ "yes" , chr.renal.D == 0 ~ "no" )) %>%
  mutate(chr.liver.D = case_when(chr.liver.D == 1 ~ "yes" , chr.liver.D == 0 ~ "no" )) %>%
  mutate(coronary.D = case_when(coronary.D == 1 ~ "yes" , coronary.D == 0 ~ "no" )) %>%
  mutate(stroke = case_when(stroke == 1 ~ "yes" , stroke == 0 ~ "no" )) %>%
  mutate(malignancy = case_when(malignancy == 1 ~ "yes" , malignancy == 0 ~ "no" ))

head(variables.of.int.recoded)

```

Let's get summary statistics with the tbl_summary() function from the "gtsummary" package. This function is handy when dealing with both numerical and categorical data: <br />
- Count and Percentage for categorical data <br />
- Median and Quartiles for numeric data <br />


```{r}
tbl_summary(variables.of.int.recoded, 
            missing = "no")
```

## Unmatched Cohorts Analysis

### 1. Crude estimate of the association between IAC and hospital mortality

Let's compute the crude estimates first using the CS() function from the EpiStats package. This function provides us with two tables: <br />
- A 2x2 contingency table illustrates the exposure and outcome of interest.<br /> 
- A table presents all possible estimates of the association.  

```{r}
result.1 <- CS(variables.of.int, death, IAC, exact = TRUE, full = FALSE)

kable(result.1$df1, align = result.1$df1.align)
kable(result.1$df2, align = result.1$df2.align)

```

Patients with an indwelling arterial catheter faced a **21%** (95%CI of -3% to 51%) to higher mortality risk compared to those without it. Additionally, **10%** of deaths can be attributed to the use of indwelling arterial catheters in ICU patients who require mechanical ventilation and are hemodynamically stable. However, this association is not significant (chi2 p = 0.085)

#### When working with multiple exposures, the CSTable() function can be quite useful providing with a single table.

Let's say sex, COPD, Other.resp.D, CHF, chr.renal.D and AF are exposures too.

```{r}
result.2 <- CSTable(variables.of.int, "death", 
        exposure = c("IAC", "sex", "COPD", "Other.resp.D", "CHF", "chr.renal.D", "AF"),
        exact = TRUE,
        sort = "rr",
        full = TRUE)
kable(result.2$df, align = result.2$align)
```

### 2. Adjusting for confounding variables using stratification

Next, we will stratify the cohort by service unit type (MICU, SICU, FICU) and compute the estimates within each stratum using the function CSInter(). 

```{r}
result.3 <- CSInter(variables.of.int, "death", "IAC", by = "service_unit", full = TRUE)

kable(result.3$df1, align = result.3$align)
kable(result.3$df2, align = result.3$align)
```
The association between IAC and mortality exists in both strata appears to be higher in the MICU, with a risk ratio of 1.46. Furthermore, the *Mantel Heaenszel risk ratio* was **1.30**	(1.04-1.64) meaning that the effect (1.21) was underestimated by the service unit type.

### 3. Multivariable model (Logistic regression) to adjust for confunding variables

Let's create our model by considering multiple confounders and generating summary and ANOVA tables using the `glm()` function. We will specify the family as "binomial" for the logistic regression, as our response variable consists of binary data (death = 1 for yes or 0 for no) and call for the summary and anova tables.

```{r}
model.1 <- glm(death ~ IAC + age + sex + sofa + service_unit + COPD + 
                 Other.resp.D + CHF + AF + chr.renal.D + chr.liver.D + 
                 coronary.D + stroke + malignancy + wbc + hemoglobin + 
                 platelet + sodium + potassium + bicarbonate + chloride + 
                 urea + creat + po2 + pco2, 
               variables.of.int, 
               family = "binomial")

summary(model.1)
anova(model.1)
```

As it can be seen, the output is a bit messy. Therefore, we will be using the tbl_regression() from the "gtsummary" package to get a tidy table.

```{r}
tbl_regression(model.1, exponentiate = TRUE)
```

Let's now get an automatic report of the model using the "report" package.

```{r, results='asis'}
report(model.1)
```

According to the model, only 9 predictor variables (age, sofa, other.resp.D, stroke, wbc, hemoglobine, bicarbonate, chloride and urea) had a significant effect on hospital mortality. If we were to reduce the model's equation to only these independent variables, the results could be as follow:

```{r}
model.2 <- glm(death ~ age + sofa + Other.resp.D + stroke + wbc + hemoglobin + bicarbonate + chloride + urea, 
               variables.of.int, 
               family = "binomial")

tbl_regression(model.2, exponentiate = TRUE)
```

#### I would like to create a summary table combining :
- all univariable logistic regressions that are possible with the variables available in our dataframe,
- a multivariable logistic regression that includes all variables available in our dataframe, and
- a multivariable logistic regression that includes only some of variables available in our dataframe.
- in addition to some descriptive statistics for each level of predictor variables.

We start with all possible univariable logistic regressions

```{r}
# Set response and predictor variables
response <- "death"
predictor <- c("IAC", "age", "sex", "sofa", "service_unit", "COPD", "Other.resp.D",
               "CHF", "AF", "chr.renal.D", "chr.liver.D", "coronary.D", "stroke",
               "malignancy", "wbc", "hemoglobin", "platelet", "sodium", "potassium",
               "bicarbonate", "chloride", "urea", "creat", "po2", "pco2")

# Save results 
glmuni <- variables.of.int %>% 
  glmuni(response, predictor) %>%
  fit2df(explanatory_name = "Variables",
         estimate_name = "Crude OR",
         estimate_suffix = " (95% CI, p value)")

kable(glmuni) 
```

Here is the multivariable logistic regression with all variables

```{r}
glmmulti_full <- variables.of.int %>% 
  glmmulti(response, predictor) %>%
  fit2df(explanatory_name = "Variables",
         estimate_name = "Adjusted OR - full model")

kable(glmmulti_full) 
```

Here is the multivariable logistic regression with all variables

```{r}
# Set predictor variables to be included in the final model
predictor_final <- c("age", "sofa", "Other.resp.D", "stroke", "wbc", "hemoglobin",
                     "bicarbonate", "chloride", "urea")

# Save results 
glmmulti_final <- variables.of.int %>% 
  glmuni(response, predictor_final) %>%
  fit2df(explanatory_name = "Variables",
         estimate_name = "Adjusted OR - final model")

kable(glmmulti_final) 
```

Let's save descriptive statistics

```{r}
summary <- variables.of.int.recoded %>% 
  summary_factorlist(response, predictor, fit_id = TRUE) 

kable(summary) 
```

Here how we can merge all those 4 tables:

```{r}
finaltab <- summary %>% 
  finalfit_merge(glmuni) %>% 
  finalfit_merge(glmmulti_full) %>% 
  finalfit_merge(glmmulti_final)

finaltab %>% select(-fit_id, -index) %>% 
  rename(Variables = label,
         " " = levels) %>% 
  kable()
```







