---
title: "Propensity Score Matching"
author: "Boaz Baliejukia"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Loading packages

```{r, message=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)
library(MatchIt)
```
# Read in data

```{r, message=FALSE}
distress <- read_csv("https://raw.githubusercontent.com/gckc123/ExampleData/main/smoking_psyc_distress.csv")

head(distress) %>% 
  kable(align = "c", format = "html") %>% 
  kable_styling(full_width = NULL, font_size = 13)
  
glimpse(distress)

```
There are ten variables in this dataset.

* Sex (0: Female; 1: Male)
* indigenous - Ingigenous status (0: Non-indigenous; 1: indigenous)
* high_school - Education level (0: not finished high school; 1: finished high school)
* partnered - Marital status (0: not partnered; 1: partnered)
* remoteness - region of residence (0: major cities; 1: inner regional; 2: outer regional)
* language - Language background (0: non-English speaking; 1: English speaking)
* Smoker - Smoking status (0: non-smoker; 1: smoker)
* risky_alcohol - Risky alcohol use (0: not risky; 1: risky)
* psyc_distress - Psychological distress. Measure ranges from 10 to 50.
* age - Age of the participants

# Data handling

The "remoteness" variable is categorical with more than 2 categories. We have to recode it as a factor

```{r}
distress$remoteness <- factor(distress$remoteness, exclude = c("", NA))
```

# Matching

```{r}
match_obj <- matchit(smoker ~ sex + indigeneity + high_school + partnered + remoteness + language + risky_alcohol + age,
                     data = distress, method = "nearest", distance = "glm",
                     ratio = 1,
                     replace = FALSE)
summary(match_obj)
```















